Neural Connectome of the Ctenophore Statocyst

### 2025-05-06 manuscript/Manuscript.qmd
Discussion

In this study, we generated a comprehensive neural map of the aboral organ of a five-day-old *Mnemiopsis leidyi*. We described a novel type of syncytial neuron and analysed its synaptic connections to balancer-cell groups across the four quadrants of the aboral organ. Given that all neurons are intrinsic to the aboral organ and our EM volume covers the entire organ, we could recover a complete circuit map, the first for a ctenophore. Contrary to our expectations, the nerve net has no motor connections to effector organs, such as the ciliary grooves or comb rows. Our high-speed imaging suggests that the nerve net instead coordinates the activity of balancer cilia. This agrees with a hydrodynamic model of graviorientation <!-- add ref to Tamm -->, with balancer cilia directly responding to statolyth load and acting as pacemakers for the comb rows via the ciliated grooves. The nerve net could in turn tune or reset sensitivites and valence of balancer cells, in a coordinatory role, as we will discuss below.

@@

### A New Type of Syncytial Neuron

We identified three syncytial neurons with multiple nuclei in the aboral organ of the cydippid larva. Burkhurdt et al. previously reported syncytial neurons in the sub-epithelial nerve net (SNN) of a one-day-old larva [@Burkhardt2023]. The syncytial neurons in the aboral organ did not exhibit the characteristic "pearl necklace-like structure" of the SNN neurons. Instead, their morphology was more reminiscent of fiber cells found in Placozoa [@Mayorova2021], with cell projections extending through intercellular spaces. In our volume, we also identified two SNN-like neurons with a pearl necklace-like morphology, indicating that the differences between ANN and SNN are not due to differences in stage or sample preparation. This confirms the ANN neurons as a second distinct type of syncytial neuron in ctenophores.

### Functional Diversification of the Aboral Syncytial Neurons

The three ANN neurons form two functional types based on their spatial arrangement and differential innervation of balancer cells.

In the this study, we found that three neurons are closely positioned within the confined space of the aboral organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner. Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency. While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.

In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Veraszto2017]. Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.

### Paracrine Signaling and Synaptic Transmission from Sensory to Neural Cells

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974]. By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the aboral organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the aboral organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the aboral organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021]. While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023]. Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells [@Hertwig1880; @Chun1880; @Eimer1880].

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the aboral organ. Bridge cells were first described by Tamm et al. based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002]. Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

### Circuit Variability as a Driver of Behavioral Diversity

Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores [@Satterlie2015a]. This finding provides evidence that increased circuit complexity contributes to behavioral diversification. By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.

The neural cells identified in the aboral organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021]. Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints [@Bullmore2012; @Perin2011]. For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring [@White1986]. Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024]. In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

@@

![**Supplemental figure**\
(A)...](figures/Supplemental_figure.png)

>>>>

### Functional Nerve Net for Ciliary Coordination

In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy. Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity.

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in Platynereis dumerilii larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands, while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits.

Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly. This departs fundamentally from reflex-arc models of early nervous systems and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.

@@

### Bridge Cells as Feedback Regulators of Ciliary Rhythms

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes, indicating spatially distinct coordination dynamics. These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia. In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.

This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells. Such a structure suggests a loop-based coordination mechanism, where the central circuit adapts its timing based on feedback from its target. Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms.

@@

![**Figure 1—figure supplement 1**\
(A)...](figures/Supplemental_figure.png)

![**Figure 2—figure supplement 1**\
(A)...](figures/Supplemental_figure.png)
